Inference of a universal social scale and segregation measures using social connectivity kernels

### 2020-09-30 manuscript/main.tex
Address referee comments.

\newenvironment{choices}
{
    \begin{itemize}
    \setlength{\itemsep}{0pt}
    \setlength{\parskip}{0pt}
    \setlength{\parsep}{0pt}
}
{\end{itemize}}

\newenvironment{questions}
{
    \begin{enumerate}
    \renewcommand{\theenumi}{\textbf{Q\arabic{enumi}}}
    \renewcommand{\labelenumi}{\theenumi:}
    \renewcommand{\labelitemi}{$\circ$}
    \setlength{\itemsep}{0pt}
    \setlength{\parskip}{0pt}
    \setlength{\parsep}{0pt}
}
{\end{enumerate}}

@@

\def\composite{}
% Useful for word counting
\ifx\composite\undefined
  % hide graphics
  \renewcommand{\includegraphics}[1]{}
  % hide pagenumber
  \pagestyle{empty}
  \renewcommand*{\titlepagestyle}{empty}
  % hide equations
  \RenewEnviron{equation}{equation environment}
  \RenewEnviron{align}{align environment}
  \RenewEnviron{equation*}{equation* environment}
\fi

@@

@@

\emph{Social} statistics of segregation, such as the assortativity coefficient~\cite{Newman2003a}, overcome these limitations by explicitly considering the interactions amongst individuals~\cite{Blumenstock2013}, but have their own difficulties: first, they usually rely on the existence of discrete groups, such as sex, ethnicity, or religion~\cite{Bojanowski2014}, and they are not applicable to continuous attributes, such as age or income. Attributes are often discretised~\cite{Lam-Morgan2012, Kalmijn2007, Kim2012}, but the boundaries between categories always suffer from some degree of arbitrariness~\cite{Reardon2004}. Second, segregation for multiple attributes can be quantified independently, but it is unclear how to define a composite segregation statistic. Furthermore, social statistics are typically defined as summary statistics of a fully-observed social network. Consequently, we cannot easily quantify uncertainties.

@@

A single statistic cannot capture the complexities of social networks, and we develop a family of statistics applicable at different scales: the social separation between any two individuals, the isolation experienced by any one individual, and the social strain experienced by society as a whole. Starting at the microscopic level, we use the generative model to define a statistic of \emph{social separation} between two individuals with attributes $x$ and $y$:
\begin{align}
    \varphi(x, y) &= \logit \rho(y,y) - \logit\rho(x,y),\label{eq:social-separation}\\
    \text{where }\logit\rho &= \log\left(\frac{\rho}{1-\rho}\right)\nonumber
\end{align}
are the log odds for a connection to form with probability $\rho$. The probability $\rho(y, y)$ for an individual with attributes $y$ to connect with someone with identical attributes serves as a reference point, and the statistic does not depend on the overall edge density. The social separation $\varphi$ may be understood as the isolation experienced by an individual with attributes $x$ as a result of the behaviour of another with attributes $y$. The statistic is zero if people do not discriminate with respect to the attributes on which they differ or if two individuals have the same demographic attributes. For a homophilous connectivity kernel, the statistic is positive and is a \emph{semi-metric} for Blau space; we will consider a family of connectivity kernels for which $\varphi$ is a true metric in \cref{sec:segregation-distance}.

@@

The connectivity kernel $\rho(x, y)$ is not sufficient to quantify segregation at the level of an individual: we also need to consider the distribution $P(y)$ of attributes $y$ of all other members of society, such as their age, sex, or other demographics. For an individual with attributes $x$, we define the \emph{social isolation}
\begin{equation}
    \phi(x) = \int dy \ P(y) \varphi(x, y),\label{eq:segregation-individual}
\end{equation}
which quantifies the average social separation between an individual with attribute $x$ and other members of society. For the SBM, we substitute \cref{eq:social-separation-sbm} into \cref{eq:segregation-individual} and obtain
\begin{equation}
    \phi(x) = (1-P(x))\left(\logit\rho_\mathrm{same}-\logit\rho_\mathrm{different}\right),\label{eq:segregation-individual-sbm}
\end{equation}
where $P(x)$ is the probability to belong to block $x$, and we have used the identity $\sum_{y=1}^K P(y)\left(1-\delta_{xy}\right)= 1 - P(x)$. Members of all blocks experience the same degree of isolation if the blocks are of the same size. If the sizes are unequal, minorities experience more isolation and majority groups experience less isolation. Indeed, ethnic minorities in schools tend to be more isolated and have fewer social ties~\cite{Currarini2009}.

@@

The social separation takes a simple form if the probability for two individuals to connect is a logistic kernel~\cite{Hoff2002}, i.e.\
\begin{equation}
    \logit\rho(x, y, \theta) = \sum_{l=1}^p\theta_l f_l(x,y),\label{eq:logistic-kernel}
\end{equation}
where the $p$-dimensional vector $\theta_l$ parametrises the kernel, and $f(x, y)$ is a set of $p$-dimensional features that are predictive of the connection probability, such as the age difference $f_\text{age}=\abs{x_\mathrm{age}-y_\mathrm{age}}$. The social separation between $x$ and $y$ comprises contributions from the features of the logistic kernel:
\begin{align}
    \varphi(x,y) &= \sum_{l=1}^p\varphi_l(x, y),\label{eq:social-separation-logistic}\\
    \text{where }\varphi_l(x, y) &= \theta_l\left(f_l(y, y) - f_l(x, y)\right)\label{eq:social-separation-contribution}
\end{align}
is the contribution due to a single feature $l$. In fact, $\varphi$ is a true metric for many logistic connectivity kernels.

@@

We considered a generative model for social networks embedded in Blau space, a space spanned by the demographic attributes of members of society. We developed a family of segregation statistics with a universal scale (since they are based on the common notion of the probability of a social tie) facilitating comparison between datasets collected at different times or in different cultural contexts. Furthermore, the segregation statistics are applicable to mixed attribute types, have a natural reference point, and an intuitive interpretation: the probability to form connections decreases with increasing segregation. For certain logistic connectivity kernels, the social separation is a metric for Blau space and allows us to quantify social distance in a principled fashion.

@@

The importance of physical distance highlights that our suite of segregation statistics does not distinguish between \emph{choice homophily} and \emph{opportunistic homophily}~\cite{Franz2010}. The former is a result of individuals having an active preference to connect with others who are alike, whereas the latter is a result of individuals being exposed to others who are similar to them. Opportunistic homophily is likely to be a large contributing factor to spatial homophily because individuals are less likely to encounter people who live far from them.

@@

The connectivity kernel is an intuitive model of how people connect with one another, and it is able to reproduce some of the statistics of real social networks. For example, people in high-density regions of Blau space have been observed to have more connections~\cite{Currarini2009}. However, exponential random graph models may be able to better capture the nature of social networks~\cite{Wimmer2010}. Furthermore, we have used a connectivity kernel that: (a)~is symmetric and cannot identify whether there is a status order in society~\cite{Chan2004,Ball2013}, and (b)~only depends on differences between individuals. For example, young men tend to have more social contacts than young women, and older women have more social contacts than older men~\cite{Bhattacharya2016}---an observation that cannot be captured by a kernel of the form we have considered. The connectivity kernel could be refined by adding the demographic attributes of the seeds and nominees as features, capturing sociability and popularity, respectively. Furthermore, it should be determined whether the number of ``intervening opportunities''~\cite{Stouffer1940}, absolute distance in Blau space, or a hybrid thereof are most predictive of tie probability. Ultimately, learning a connectivity kernel without a pre-specified parametric form should be considered~\cite{Frolich2006}. We note that, irrespective of the choice of connectivity kernel, the interpretable segregation statistics considered here remain valid and useful.

@@

We would like to thank Sahil Loomba for useful feedback on the manuscript and acknowledge support from the grant EP/N014529/1.

@@

\ifdefined\composite
\pagebreak
\noindent{\titlefont\huge Supplementary material}

\appendix
\numberwithin{equation}{section}
\numberwithin{figure}{section}
\numberwithin{table}{section}

>>>>

@@

% Command to highlight changes
\newcommand{\change}[1]{\textbf{#1}}

@@

\def\composite{composite}

% Useful changes for word counting
% \def\mywordcount{wordcount}
\ifdefined\mywordcount
  % hide graphics
  \renewcommand{\includegraphics}[1]{}
  % hide pagenumber
  \pagestyle{empty}
  \renewcommand*{\titlepagestyle}{empty}
  % hide equations
  \RenewEnviron{equation}{}
  \RenewEnviron{align}{}
  \RenewEnviron{equation*}{}
  \RenewEnviron{multline}{}
  % hide inline equations
  \def\[#1\]{}
  \def\(#1\){}
  \catcode`\$=13
  \def$#1${}
  % hide section numbers
  % \renewcommand{\section}[1]{\section*{#1}}
  % \setcounter{secnumdepth}{0}
\fi

@@

\emph{Social} statistics of segregation, such as the assortativity coefficient~\cite{Newman2003a}, overcome these limitations by explicitly considering the interactions amongst individuals~\cite{Blumenstock2013}, but have their own difficulties: first, they usually rely on the existence of discrete groups, such as sex, ethnicity, or religion~\cite{Bojanowski2014}, and they are not applicable to continuous attributes, such as age or income. Attributes are often discretised~\cite{Lam-Morgan2012, Kalmijn2007, Kim2012}, but the boundaries between categories always suffer from some degree of arbitrariness~\cite{Reardon2004}. Second, segregation for multiple attributes can be quantified independently, \change{but \textcite{Pelechrinis2016}, who consider a multivariate generalisation of the assortativity coefficient, note that ``a formal metric that is generally applicable'' remains elusive.} Furthermore, social statistics are typically defined as summary statistics of a fully-observed social network. Consequently, we cannot easily quantify uncertainties.

@@

A single statistic cannot capture the complexities of social networks, and we develop a family of statistics applicable at different scales: \change{(a)~the \emph{social separation} between any two individuals, (b)~the \emph{isolation} experienced by any one individual, and (c)~the \emph{social strain} experienced by society as a whole.} Starting at the microscopic level, \change{we define the \emph{social separation} between two individuals $i$ and $j$ with attributes $x$ and $y$ as the relative log odds for $j$ to connect with $i$ compared someone who is alike: the log odds intuitively capture the probabilistic nature of the conditionally independent edge model. In particular,}
\begin{align}
    \varphi(x, y) &= \logit \rho(y,y) - \logit\rho(x,y),\label{eq:social-separation}\\
    \text{where }\logit\rho &= \log\left(\frac{\rho}{1-\rho}\right)\nonumber
\end{align}
are the log odds for a connection to form with probability $\rho$\change{~\cite{Hastie2009}}. The probability $\rho(y, y)$ for \change{$j$ to connect with someone who is alike} serves as a reference point, and the statistic does not depend on the overall edge density. The social separation $\varphi$ may be understood as the isolation experienced by \change{$i$ with attributes $x$ as a result of the behaviour of $j$ with attributes $y$}. The statistic is zero if people do not discriminate with respect to the attributes on which they differ or if two individuals have the same demographic attributes. For a homophilous connectivity kernel, the statistic is positive and is a \emph{semi-metric} for Blau space; we will consider a family of connectivity kernels for which $\varphi$ is a true metric in \cref{sec:segregation-distance}.

@@

The connectivity kernel $\rho(x, y)$ is not sufficient to quantify segregation at the level of an individual: we also need to consider the distribution $P(y)$ of attributes $y$ of all other members of society, such as their age, sex, or other demographics. For an individual with attributes $x$, we define the \emph{social isolation}
\begin{equation}
    \phi(x) = \int dy \ P(y) \varphi(x, y),\label{eq:segregation-individual}
\end{equation}
which quantifies the average social separation between an individual with attribute $x$ and other members of society. For the SBM, we substitute \cref{eq:social-separation-sbm} into \cref{eq:segregation-individual} and obtain
\begin{equation}
    \phi(x) = (1-P(x))\left(\logit\rho_\mathrm{same}-\logit\rho_\mathrm{different}\right),\label{eq:segregation-individual-sbm}
\end{equation}
where $P(x)$ is the probability to belong to block $x$, and we have used the identity $\sum_{y=1}^K P(y)\left(1-\delta_{xy}\right)= 1 - P(x)$. Members of all blocks experience the same degree of isolation if the blocks are of the same size. If the sizes are unequal, minorities experience more isolation and majority groups experience less isolation. Indeed, ethnic minorities in schools tend to be more isolated and have fewer social ties~\cite{Currarini2009}. \change{The off-diagonal terms of the Hessian of the social isolation $\phi$ quantify interactions, such as the joint effect of age and ethnic differences.}

@@

The social separation takes a simple form if the probability for two individuals to connect is a logistic kernel~\cite{Hoff2002}, i.e.\
\begin{equation}
    \logit\rho(x, y, \theta) = \sum_{l=1}^p\theta_l f_l(x,y),\label{eq:logistic-kernel}
\end{equation}
where the $p$-dimensional vector $\theta_l$ parametrises the kernel, and $f(x, y)$ is a set of $p$-dimensional features that are predictive of the connection probability, such as the age difference $f_\text{age}=\abs{x_\mathrm{age}-y_\mathrm{age}}$. \change{Intersectionality can be accounted for by including interaction terms in the feature set.} The social separation between $x$ and $y$ comprises contributions from the features of the logistic kernel:
\begin{align}
    \varphi(x,y) &= \sum_{l=1}^p\varphi_l(x, y),\label{eq:social-separation-logistic}\\
    \text{where }\varphi_l(x, y) &= \theta_l\left(f_l(y, y) - f_l(x, y)\right)\label{eq:social-separation-contribution}
\end{align}
is the contribution due to a single feature $l$. In fact, $\varphi$ is a true metric for many logistic connectivity kernels.

@@

We considered a generative model for social networks embedded in Blau space, a space spanned by the demographic attributes of members of society. We developed a family of segregation statistics with a universal scale (since they are based on the common notion of the probability of a social tie), facilitating comparison between datasets collected at different times or in different cultural contexts. Furthermore, the segregation statistics are applicable to mixed attribute types, have a natural reference point, and an intuitive interpretation: the probability to form connections decreases with increasing segregation. \change{They are applicable at different resolutions: connections, individuals, and society as a whole.} For certain logistic connectivity kernels, the social separation is a metric for Blau space and allows us to quantify social distance in a principled fashion. \change{The model-based approach facilitates the study of segregation in synthetic social networks, the effect of interventions, and principled quantification of uncertainties in an applied setting.}

@@

The importance of physical distance highlights that our suite of segregation statistics does not distinguish between \emph{choice homophily} and \emph{opportunistic homophily}~\cite{Franz2010}. The former is a result of individuals having an active preference to connect with others who are alike, whereas the latter is a result of individuals being exposed to others who are similar to them. Opportunistic homophily is likely to be a large contributing factor to spatial homophily because individuals are less likely to encounter people who live far from them. \change{Similarly, the statistics do not discriminate between choice homophily and \emph{social influence}, i.e.\ the tendency for people to become more alike given a connection~\cite{Shalizi2011}.}

\change{Other features, including sex, ethnicity, religion, education, and occupation, have smaller effects on the presence of connections. Notably, the social strain due to ethnicity in the GSS and ALP is larger than in the USS: first, the effect of ethnicity is more pronounced in the US, as shown in \cref{fig:parameters-and-age-equivalents}. Second, society in the US is more ethnically diverse than in the UK (79\% white in GSS '04 compared with 89\% white in USS '14), increasing strain on average, as exemplified with a SBM in \cref{sec:model-based-segregation}.}

@@

The connectivity kernel is an intuitive model of how people connect with one another, and it is able to reproduce some of the statistics of real social networks. For example, people in high-density regions of Blau space have been observed to have more connections~\cite{Currarini2009}. However, exponential random graph models may be able to better capture the nature of social networks~\cite{Wimmer2010}. Furthermore, we have used a connectivity kernel that: (a)~is symmetric and cannot identify whether there is a status order in society~\cite{Chan2004,Ball2013}, and (b)~only depends on differences between individuals. For example, young men tend to have more social contacts than young women, and older women have more social contacts than older men~\cite{Bhattacharya2016}---an observation that cannot be captured by a kernel of the form we have considered. The connectivity kernel could be refined by adding the demographic attributes of the seeds and nominees as features, capturing sociability and popularity, respectively. Furthermore, it should be determined whether the number of ``intervening opportunities''~\cite{Stouffer1940}, absolute distance in Blau space, or a hybrid thereof are most predictive of tie probability. Ultimately, learning a connectivity kernel without a pre-specified parametric form should be considered~\cite{Frolich2006} \change{because they can better capture complex patterns, such as interactions between different demographic attributes}. We note that, irrespective of the choice of connectivity kernel, the interpretable segregation statistics considered here remain valid and useful.

@@

We would like to thank Sahil Loomba \change{and two anonymous referees} for useful feedback on the manuscript and acknowledge support from the grant EP/N014529/1.

@@

\ifdefined\composite
\pagebreak
\setcounter{page}{1}
\noindent{\titlefont\huge Supplementary material}

\appendix
\vspace{1em}\noindent\change{The code to reproduce the results and figures is available at \url{https://github.com/tillahoffmann/kernels}.}
\numberwithin{equation}{section}
\numberwithin{figure}{section}
\numberwithin{table}{section}
